---
title: 'Stat 272: FINAL PROJECT'
subtitle: 'STAGE 1: PROPOSAL (Emelyce & Inti & Nancy)'
output:
  pdf_document:
    fig_height: 3
    fig_width: 4.5
  html_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
---


```{r, include= FALSE}
#loading necessary libraries
library(tidyverse)
library(readr)
```


# Stage I: Proposal & Data Assembly

# DATA & DATA SOURCE.

We used the UCI Machine Learning Repository as our data source: \href{https://archive.ics.uci.edu/dataset/17/breast+cancer+wisconsin+diagno}{Link.}

```{r}
#Load Dataset!
breast_cancer_data <- read_csv("~/Stats 272 S24/Project/Emelyce_Inti_Nancy /clean_breast_cancer_data.csv")

head(breast_cancer_data)

```

# RESEARCH QUESTION: What features of breast tumor cells are associated with a cell being malignant or benign? 


As the medical industry is yet to find a cure for cancer, the possibility of early detection of cancer cells or cancerous tumors can save lives. The use of statistical modeling to predict potential cancerous cells/tumors has gained a lot of momentum recently as the availability of computational power and relevant data increased exponentially over the past decade. We decided to pursue the above research question due to its relevance and potential implications in our healthcare system. We are planning to use the following 11 variables for the modeling:

Ten real-valued features are computed for each cell nucleus from breast tumors:
\begin{itemize}
    \item Diagnosis (\textit{M} = malignant, \textit{B} = benign)
    \item radius (mean of distances from center to points on the perimeter)
    \item texture (standard deviation of gray-scale values)
    \item perimeter
    \item area
    \item smoothness (local variation in radius lengths)
    \item compactness (perimeter$^2$ / area - 1.0)
    \item concavity (severity of concave portions of the contour)
    \item concave points (number of concave portions of the contour)
    \item symmetry
    \item fractal dimension ("coastline approximation" - 1)
\end{itemize}

These variables are used to create the model due to precedence in previous breast cancer research such as \href{https://archive.ics.uci.edu/dataset/17/breast+cancer+wisconsin+diagno}{Pan, Jian et al. “Bagging-Based Logistic Regression With Spark: A Medical Data Mining Method.” (2016)}.


# REFERENCES 

1. Pan, J., Yiang, H., Liu, X., Chen, Z., & Yan, Z. (2016). Bagging-Based Logistic Regression With Spark: A Medical Data Mining Method.

**Rationale**:This article presents an efficient approach for analyzing big medical data with the need for advanced techniques in healthcare research. Given our research question on features associated with breast tumor malignancy, the proposed bagging-based logistic regression method, optimized for parallel computing. The method used details that researchers can potentially enhance the accuracy of predicting tumor behavior based on features such as radius, texture, and perimeter, thus contributing to early detection of cancers and improved healthcare outcomes.


2. Street, W.N., Wolberg, W.H., & Mangasarian, O.L. (1993). Nuclear feature extraction for breast tumor diagnosis. Electronic imaging.

**Rationale**:The article utilizes an interactive image processing and statistical modeling for accurate breast tumor diagnosis. With the extraction of ten crucial features from cell nuclei, including texture, area, and smoothness, the article achieved an impressive 97% accuracy in distinguishing between benign and malignant tumors. This approach aligns closely with our research on identifying features associated with tumor malignancy, suggesting potential relevance in improving  predictive modeling for early cancer detection and prognosis assessment.



# VARIABLE CHART

The units of the variables are not provided in the original paper and the data source and for our roject they are not relevant. 

```{r}
library(kableExtra)

# Create a data frame with the variable information
variable_info <- data.frame(
  Name = c("Diagnosis", "Radius", "Texture", "Perimeter", "Area", "Smoothness", "Compactness", "Concavity", "Concave points", "Symmetry", "Fractal dimension"),
  Variable_Role = c("Response", "Explanatory", "Explanatory", "Explanatory", "Explanatory", "Explanatory", "Explanatory", "Explanatory", "Explanatory", "Explanatory", "Explanatory"),
  Type = c("Categorical", "Quantitative", "Quantitative", "Quantitative", "Quantitative", "Quantitative", "Quantitative", "Quantitative", "Quantitative", "Quantitative", "Quantitative"),
  Values = c("M/B", "6.981 - 28.110", "9.71 - 39.28", "43.79 - 188.50", "143.5 - 2501.0", "0.0526 - 0.163", "0.0194 - 0.345", "0.000 - 0.427", "0 - 0.201", "0.106 - 0.304", "0.0499 - 0.0974"),
  Units = c("NA", "", "", "", "", "", "", "", "", "", "")
)


# Print the table
kable(variable_info, "latex", booktabs = TRUE) %>%
  kable_styling(full_width = FALSE)
```


# HOW WE PLAN TO ADDRESS RESEARCH QUESTION: 

We plan to use logistic regression models with Diagnosis (M = malignant, B = benign) as response and use the rest of the features of the cells and possibly their interactions as explanatory variables. 


~END!

