---
title: "MSCS 264: Homework #4"
subtitle: "Due Fri, Feb 24 at 11pm"
author: "AL Ashir Intisar"
output:
  pdf_document:
    fig_height: 3
    fig_width: 4.5
  html_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
---

This homework focuses on mutate and using pipes to connect multiple data transformations together.


```{r, message = FALSE}
library(tidyverse)
library(lubridate)
library(nycflights13)
```



One of the advantages of pipes is that it makes code very readable. For example, this chunk of code

```{r, eval = FALSE}
mhealth_full %>%
  filter(Group == "By Age") %>%
  rename(date = `Time Period Start Date`) %>%
  select(date, Subgroup, Value)
```
can be read as "Start with mhealth_full dataset, then filter to include only Group "By Age", then rename the Time Period Start Date variable, then select only the variables date, subgroup and value."


1. Consider our diamonds dataset. We usually start with code that looks similar to this. Write a sentence that states what this code does (similar to the style above). (Hint: use ?slice_sample to check out help menu).

```{r}
smaller <- diamonds %>%
  filter(carat <= 3) %>%
  slice_sample(n = 1000)
```

**Ans: slice_sample() function randomly selects rows from the dataset provided where n = number of rows to be selected. The pipeline takles in the dataset diamonds and then filters out the records (rows) with more than 3 carat diamonds and then takes a random 1000 row sample of the filtered data.**


2. Your friend wrote the following chunks of code to create a plot. 
Use pipes to connect from "diamonds" in #1 above, all the way through the plot. Modify the order of steps so that your graph includes the prices for 1000 diamonds that are 1/2 carat or less.
```{r}
smaller_graph <- diamonds |>
  filter(carat <= 0.5) |>
  slice_sample(n = 1000)|>
  ggplot() +
    geom_boxplot(aes(y = color, x = price)) + 
    labs(y = "Color", x = "Price (USD)", title = "Diamond Color and Price, Diamonds less than 1/2 carat only")
```

```{r}
smaller_graph
```


3. Using the diamonds dataset, complete the following steps: 

- restrict to diamonds between 1 and 2 carats (include diamonds exactly 1 carat, but not exactly 2 carats) and color D, E, or F.

```{r}
diamonds_3 <- diamonds|>
  filter(carat >=1, carat < 2, color == c("D", "E", "F"))
```

```{r}
head(diamonds_3)
```


- create a variable that indicates if the diamond costs more than $10000.

```{r}
diamonds_3 <- diamonds_3|>
  mutate("Cost > $10,000" = price>10000)|>
  arrange(desc(price))
```

```{r}
head(diamonds_3)
```


- create a variable that indicates if a diamond is in one of the top clarity categories (hint: high_clar = ifelse(clarity %in% c("VVS2", "VVS1", "IF"), "high clarity", "low clarity"))

```{r}
diamonds_3 <- diamonds_3|>
  mutate("high clar" = high_clar <- ifelse(diamonds_3$clarity %in% c("VVS2", "VVS1", "IF"), "high clar", "low clar"))
```

```{r}
head(diamonds_3)
```


- create a bar chart that looks like "hw4_clarity_cost_color.png" in the homework images folder.

```{r}
diamonds_3|>
  ggplot()+
  geom_bar(mapping = aes(`high clar`, fill = `Cost > $10,000`), position = "fill")+
  facet_wrap(~color)+
  xlab("High Clarity = VVS2, VVS1, or IF rating")
```


For our next problem, we use the basketball data again.
```{r setup, include = FALSE}
library(rvest)

url = glue::glue("http://www.basketball-reference.com/leagues/NBA_2022_totals.html")

bball = read_html(url) %>% 
  html_nodes("#totals_stats") %>% 
  html_table() %>%
  data.frame(check.names = FALSE) %>%
  as_tibble() %>%
  mutate(across(G:PTS, parse_number),
         points_per_minute = PTS/MP,
         Pos = str_sub(Pos, 1,2),
         Pos = str_replace(Pos, "-", "")) %>%
  rename(FGpct = `FG%`,
         FTpct = `FT%`) %>%
  filter(MP > 10)
  

bball
```

4. In the previous homework, we found the top scoring centers using the following steps.  Rewrite this code using pipes. Also add a line using slice_max to print the top 3 scoring centers.


```{r}
bball3 <- bball|>
  filter(Pos == "C")|>
  select(Player, Tm, PTS)|>
  arrange(desc(PTS))|>
  rename(Team = Tm, "Points Scored" = PTS)

slice_max(bball3, `Points Scored`, n = 3)
```


### Map US trends in COVID vaccinations

In this portion, we are going to replicate the plot of covid vaccinations found here: [CDC webpage for tracking COVID vaccinations](https://covid.cdc.gov/covid-data-tracker/#vaccination-trends). 
(Note that our dataset only goes through 2022, while the plot on the website goes through 2023.)


Read the data into R using the chunk below. 

```{r, message = FALSE}
library(tidyverse)
library(lubridate)
vaccinations <- read_csv("~/Mscs 264 S23/Class/Data/vaccinations.csv")
vaccinations
```

5. Using pipes, apply the following transformations to the vaccinations dataset. (Hint: do them one at a time, checking each time to see if it worked, then pipe to the next one.)
    - `rename` the variables "Total Doses Administered Daily" to "total_daily" and "7-Day Avg Total Doses Daily" to "total_7day_avg"
    - `select` only the relevant columns: Date, total_daily, total_7day_avg
    - `mutate` new variaables, total_daily_millions and total_7day_avg_millions by dividing by 1000000.



```{r}
vaccinations <- vaccinations|>
  rename(total_daily = `Total Doses Administered Daily`, total_7day_avg = `7-Day Avg Total Doses Daily`)|>
  select(Date, total_daily, total_7day_avg)|>
  mutate(total_daily_millions = total_daily/1000000, total_7day_avg_millions = total_7day_avg/1000000)
```

```{r}
head(vaccinations)
```

    

6. Also create "Date = mdy(Date)" using mutate. What does this do? (Hint: Print the dataset to the console before and after adding this mutate code.)

```{r}
vaccinations <- vaccinations|>
  mutate(Date = mdy(Date))
```

```{r}
head(vaccinations)
```

**Ans: The function mdy() reverses the order of how the date is represented. NOw it is presented as Year-Month-Date**

7. Your resulting dataset should look like vax_tidy (see below). (You can use your dataset from above, or just use vax_tidy if you're not confident about #5-6). I've put code to start your plot here.  Add the line for 7-day average to the plot using geom_line. You can see a list of colors available [here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf), and experiment to get the width of the line to look nice.


```{r}
vax_tidy <- read_csv("~/Mscs 264 S23/Class/Data/vax_tidy.csv")
vax_tidy
```


```{r}
ggplot(vaccinations) +
  geom_col(aes(Date, total_daily_millions), color = "black", size = .05)+
  geom_line(aes(Date, total_7day_avg_millions), color = "purple",  size = 2)
```


8. Add labs to your plot for the y axis and a title. See "hw4_vax_plot.png" in the Homework > images folder as an example of what your resulting plot should look like. 

```{r}
ggplot(vaccinations) +
  geom_col(aes(Date, total_daily_millions), color = "black", size = .05)+
  geom_line(aes(Date, total_7day_avg_millions), color = "purple",  size = 2)+
  labs(title = "Daily Count of Total Doses Administered and Reported")+
  xlab("Date")+
  ylab("Doses (millions)")
```

