---
title: "HW 9"
author: "Your Name Here"
date: "3/9/2023"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

## Part 1: data drawings

This is our dataset. It is named `data`

(Note: I used a random name generator on the internet, and then randomly assigned favorite foods, major, and number of credits this semester!)

Note that the variables Name and Fav_food are character variables. 
Major is a factor variable (<fct>). The numbers in parentheses indicate the order. 
So, if we were actually looking at this data in R, it would just print that Nia's major is "art", not "art(1)"... but the level order 1 is stored internally.


Name    Fav_food Major    Credits
------- -------- -------- --------
Nia     ramen    art(1)   3.5
Lionel  pizza    chem(3)  4.0
Yan     ramen    art(1)   4.5
Jie     pizza    art(1)   3.0
Tom     pizza    bio(2)   4.0
Cloe    tacos    chem(3)  3.0   
Carlo   pizza    bio(2)   4.0
Ping    tacos    dance(4) 3.5



For the following problems, copy the table above, and then modify it according to each instruction. 

For example:

What does the dataset look like after running this code?
```
data %>%
  select(Name, Fav_food)
```
Answer: 

Name   Fav_food  
------ ---------
Nia     ramen    
Lionel  pizza    
Yan     ramen    
Jie     pizza    
Tom     pizza   
Cloe    tacos    
Carlo   pizza    
Ping    tacos    



1. What does the dataset look like after running this code?
```
data %>%
  filter(Fav_food == "ramen")
```


Answer:


Name    Fav_food Major    Credits
------- -------- -------- --------
Nia     ramen    art       3.5
Yan     ramen    art       4.5

2. What does the dataset look like after running this code? 
```
data %>%
  count(Major)
```

Answer: 

Major   n  
------ ---------
art     3    
bio     2    
chem    2    
dance   1    

3. What does the dataset look like after running this code?
```
data %>%
  arrange(Name)
```

Answer:

Name    Fav_food Major    Credits
------- -------- -------- --------
Carlo   pizza    bio      4.0
Cloe    tacos    chem     3.0
Jie     pizza    art      3.0
Lionel  pizza    chem     4.0
Nia     ramen    art      3.5
Ping    tacos    dance    3.5
Tom     pizza    bio      4.0
Yan     ramen    art      4.5

   



3. What does the dataset look like after running this code?
```
data %>%
  mutate(cred4plus = Credits >=4)
```

Answer:

Name    Fav_food Major    Credits   cred4plus
------- -------- -------- --------  ---------
Nia     ramen    art      3.5       FALSE
Lionel  pizza    chem     4.0       TRUE
Yan     ramen    art      4.5       TRUE
Jie     pizza    art      3.0       FALSE
Tom     pizza    bio      4.0       TRUE
Cloe    tacos    chem     3.0       FALSE
Carlo   pizza    bio      4.0       TRUE
Ping    tacos    dance    3.5       FALSE


4. What does the dataset look like after running this code?
```
data %>%
  rename(credits_spring2023 = Credits)
```

Answer:

Name    Fav_food Major    credits_spring2023
------- -------- -------- -------------------
Nia     ramen    art      3.5
Lionel  pizza    chem     4.0
Yan     ramen    art      4.5
Jie     pizza    art      3.0
Tom     pizza    bio      4.0
Cloe    tacos    chem     3.0   
Carlo   pizza    bio      4.0
Ping    tacos    dance    3.5

4. What does the dataset look like after running this code?
```
data %>%
  group_by(Major) %>%
  summarize(max_credits = max(Credits))
```

Answer: 

Major   max_credits  
------  ------------
art     4.5    
bio     4.0    
chem    4.0    
dance   3.5

5. What does the dataset look like after running this code? 
```
data %>%
  mutate(Major = fct_reorder(Major, Credits, .fun = max))
```

Answer:

Name    Fav_food Major    Credits
------- -------- -------- --------
Nia     ramen    art(1)   3.5
Lionel  pizza    chem(3)  4.0
Yan     ramen    art(1)   4.5
Jie     pizza    art(1)   3.0
Tom     pizza    bio(2)   4.0
Cloe    tacos    chem(3)  3.0   
Carlo   pizza    bio(2)   4.0
Ping    tacos    dance(4) 3.5

# Part 2: Spotify data
```{r, message = FALSE}
library(tidyverse)
```


```{r}
bigspotify <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-21/spotify_songs.csv') %>%
  mutate(year = parse_number(substr(track_album_release_date, 3, 4)),
         decade = (year %/% 10)*10,
         decade = str_c(as.character(decade), "s", sep = ""))

bigspotify %>% count(decade)
```

6. Above you can see that I've tried to create a variable to indicate the decade of a song... but it needs some help. First, modify the decade variable so that the aughts are "00s" instead of "0s". Print your result with count (see below):

```{r}
bigspotify|>
  mutate(decade = as.factor(decade))|>
  mutate(decade = fct_recode(decade, "00s" = "0s"))|>
  count(decade)  
```


7. Now add another line to the code above so that 00s, 10s, and 20s, come AFTER the 90s. Again, print your result with count.

```{r}
bigspotify|>
  mutate(decade = as.factor(decade))|>
  mutate(decade = fct_recode(decade, "00s" = "0s"))|>
  mutate(decade = fct_relevel(decade, "90s"))|>
  count(decade) 
```



8. Find the average track_popularity, danceability, and speechiness by decade. Write 1 sentence about what you observe for each variable. (If you got 6-7 to work, use that ordering for decade. If not, just use the original bigspotify!)

```{r}
bigspotify|>
  mutate(decade = as.factor(decade))|>
  mutate(decade = fct_recode(decade, "00s" = "0s"))|>
  mutate(decade = fct_relevel(decade, "90s"))|>
  group_by(decade)|>
  summarise(mean_pop = mean(track_popularity), mean_dance = mean(danceability), mean_speech = mean(speechiness))

```


9. Create a plot that helps to visualize speechiness by playlist_genre and decade. Be sure that playlist_genre is in a meaningful order in your graph.
Write 1-2 sentences about what you observe. (Note: you might want to drop the 50s since there are so few songs!)

```{r}
bigspotify|>
  mutate(playlist_genre = fct_reorder(playlist_genre, speechiness, .fun = mean))|>
  group_by(playlist_genre, decade)|>
  summarise(mean_speechiness = mean(speechiness))|>
  ggplot(aes(playlist_genre, decade))+
  geom_point(aes(size = mean_speechiness, color = decade))
```

**Ans: Rock has the lowest mean speechiness and the rap genre has the highest mean speechiness. Also the speechiness in rock genre is more evenly distributed in rock and r&b than other genres. There is not much change in speechiness with respect to changes in decades for a specific genre.**  

10. Pick a new theme and/or color scale for your plot above! 
(hint start typing theme_ then hit tab to see options!)

```{r}
bigspotify|>
  mutate(playlist_genre = fct_reorder(playlist_genre, speechiness, .fun = mean))|>
  group_by(playlist_genre, decade)|>
  summarise(mean_speechiness = mean(speechiness))|>
  ggplot(aes(playlist_genre, decade))+
  geom_point(aes(size = mean_speechiness, color = decade))+
  theme_dark()
```


