---
title: "MSCS 264: Homework #12"
subtitle: "Due Tues, April 11 at 11pm"
output:
  pdf_document:
    fig_height: 3.5
    fig_width: 6
  word_document: default
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
author: Al Ashir Intisar
---

As always, submit your knitted pdf file to Moodle, but be sure your RMarkdown file is saved and accessible in your Submit folder on the RStudio server.

```{r, include = FALSE, message=FALSE}
library(tidyverse)
library(rvest)
library(ggplot2)
data(mpg)
data(gss_cat)
```

# A summary stats function

1. In this problem, we we will create a function called `summary_stats()` to produce a custom set of summary statistics (n, mean, median, sd, IQR, min, and max) for any variable that you input from a tibble. Follow the steps below to create your function.

(a) First, lets start with working code. Use `summarize` to calculate n, mean, median, sd, IQR, min, and max, for the hwy variable in the mpg dataset. 

```{r}
mpg|>
  summarise(n = n(), mean = mean(hwy), median = median(hwy), IQR = IQR(hwy), min = min(hwy), max = max(hwy))
```



(b) Copy your code from above and make it into a function with the name `summary_stats()`. It should have two inputs, `data` (the tibble of interest) and `x` (the variable of interest). 

```{r}
summary_stats <- function(tbl, variable){
  tbl|>
    summarise(n = n(), mean = mean({{variable}}), median = median({{variable}}), IQR = IQR({{variable}}), min = min({{variable}}), max = max({{variable}}))
    
}
```


    Hint: because of ["tidy evaluation"](https://dplyr.tidyverse.org/articles/programming.html), when you reference a variable from a tibble inside a function, you have to "embrace" the variable name in double brackets; e.g. `{{ x }}`, where `x` is the variable name.
    
    

    
(c) Test your function on the hwy and cty variables from mpg.

```{r}
summary_stats(mpg, hwy)
summary_stats(mpg, cty)
```



(d) Test your function on the age variable from gss_cat. Why do you get this output?

```{r, error=TRUE}
summary_stats(gss_cat, age)
```



(e) Add an option to remove missing values, if any exist (and be sure the sample size `n` reflects the number of non-missing values).  Thus, your function should have 3 inputs: `data` (the tibble of interest), `x` (the variable of interest), and `removeNA` (with a default value of FALSE).  

```{r}
summary_stats <- function(tbl, variable, removeNA = FALSE){
  
  if(removeNA == FALSE){
    tbl|>
    summarise(n = n(), mean = mean({{variable}}), median = median({{variable}}), IQR = IQR({{variable}}), min = min({{variable}}), max = max({{variable}}))
    
  }
  else{
     tbl|>
      drop_na()|>
    summarise(n = n(), mean = mean({{variable}}), median = median({{variable}}), IQR = IQR({{variable}}), min = min({{variable}}), max = max({{variable}}))
    
  }
  
    
}
```



(d) Test your new function on the hwy and cty variables from mpg, and the age variable from gss_cat.

```{r}
summary_stats(mpg, hwy, TRUE)
summary_stats(mpg, cty, TRUE)
summary_stats(gss_cat, age, TRUE)
```



# Scraping movie data from IMDB

2. Go to the following website, and describe what the page shows: 
https://www.imdb.com/search/title?year=2017&title_type=feature&sort=boxoffice_gross_us,desc


**Ans: The website page rates movies released in the year 2017 and also includes other data like title, genre, r rating status, number of revies, gross income from the movie etc. It is basically a movie rating web page.**


3. The code below pulls the title and box office gross (money earned by tickets) for the top 50 grossing movies of a given year from the webpage! Look at "gross", "titles" and the resulting tibble! Turn this code into a function called "scrape_top50" with an input "movie_year". (Look closely! anywhere else that refers to 2017?)


```{r}
scrape_top50 <- function(movie_year){
  
url <- str_c("https://www.imdb.com/search/title?year=", movie_year, "&title_type=feature&sort=boxoffice_gross_us,desc")

# Don't worry about being able to write this code! We'll get there!
gross <- read_html(url) %>%
    html_nodes(".ghost~ .text-muted+ span") %>%
  html_text()

titles <- read_html(url) %>%
  html_nodes(".lister-item-header a") %>%
  html_text()

# This part you should understand!
tibble(movie = titles,
       gross = gross,
       year = movie_year)
  
}


```


4. Now that you have written your function, you can scrape data for multiple years! Scrape the data for 2015 to 2019. Name each resulting tibble "movies_2015" etc.  Then you can use this code to put them together into one dataset: 

```{r, error = TRUE}
movies_2015 <- scrape_top50(2015)
movies_2016 <- scrape_top50(2016)
movies_2017 <- scrape_top50(2017)
movies_2018 <- scrape_top50(2018)
movies_2019 <- scrape_top50(2019)

all_movies <- movies_2015 %>%
  bind_rows(movies_2016) %>%
  bind_rows(movies_2017) %>%
  bind_rows(movies_2018) %>%
  bind_rows(movies_2019)
```


5. If you weren't able to get all_movies, you can read in the datset from our class code folder! Use either version to: 

a) create a numeric variable called "gross_millions", which is the gross in millions of US dollars.

```{r}
all_movies <- all_movies|>
  mutate(gross_millions = parse_number(gross))
```


b) Use your summary_stats function to calculate a table of summary stats of "gross_millions" for each year. 
(Note: If you didn't get `summary_stats` working above, just calculate the mean gross_millions for each year).

```{r}
all_movies|>
  group_by(year)|>
  summary_stats(gross_millions, TRUE)
```



c) From all_movies, get the top grossing movie for each year. Create a graph with year on the x axis, gross_millions on the y axis. Label each point with the title of the movie, and connect them with a line. Give your graph an appropriate title and axis labels. Use a new color and/or theme!
Also use   `coord_cartesian(ylim = c(0, 1000))` to make the y-axis go from 0 to 1 billion. 


```{r}
library(ggrepel)
all_movies|>
  group_by(year)|>
  slice_max(gross_millions, n = 1)|>
  ggplot(aes(x = year, y = gross_millions))+
  geom_point()+
  geom_line()+
  #geom_label(aes(label = movie))+
  geom_text_repel(aes(label = movie), nudge_y = -30, nudge_x = .3, segment.size = 0) +  # Add labels
  #geom_segment(aes(x = year, y = gross_millions, xend = year + 0.2, yend = gross_millions - 25), 
               #arrow = arrow(length = unit(0.15, "cm")), show.legend = FALSE)+
  coord_cartesian(ylim = c(0, 1000))+
  labs(title = "Maximum Gross earning movie of the year from year 2015-2019", x = "Year of release", y = "Gross earning in millions $")+
  theme_minimal()
  
```


d) Write the all_movies dataset to your submit folder!

```{r}
write.csv(all_movies, "~/Mscs 264 S23/Submit Section A/Homeworks /all_movies.csv")
```


